Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-18 paper.tex
progress

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}

@@

\begin{enumerate}
\item In Section \ref{sec:ttgu} we describe models of type theories where
  universe levels may come from any set with a well-founded transitive ordering
  relation. We specify models as categories equipped with level-indexed diagrams
  of families, as a variation on categories with families. Each morphism of
  levels is mapped to a lifting operation on terms and types. By varying the
  preservation properties of lifting operations, we can describe a range of
  stratification features, from two-level type theory to cumulative universes.
\item We use induction-recursion to model the mentioned theories. We model
  the strongest formulations for lifting and universes, namely cumulative
  universes with Russell-style type decoding. The main challenge is combining
  transfinite hierarchies with strictly type-preserving lifting operations.
\item In Section \ref{sec:ttfl} we describe type theories with internal types
  for levels and level morphisms, and extend the previous inductive-recursive
  semantics to cover these as well. Here, we can additionally represent various
  universe polymorphism features and level computations. We illustrate how
  predicative universe features in Coq and Agda could be justified in our framework.
\end{enumerate}

@@

\section{Type Theories with Generalized Universe Hierarchies}
\label{sec:ttgu}

@@

\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$ all
descending $R$-chains must be finite. This is ensured by the universal property
of the inductive definition, as all inductive types must be well-founded.

\begin{mylemma}
Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
propositionally equal \cite[Lemma 10.3.4]{hottbook}.
\end{mylemma}

@@

Unlike every other type or term former, there is no lifting computation rule for
$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
lifting and term lifting, but while term lifting has invertible action, type
lifting does not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a
larger $\Ty$ is not isomorphic a smaller one, because it contains more
universes. So, for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to
$\Ty_1\,\Gamma$ yields $\Bool_1$, but lifting $\Bool_0$ as a term yields
$\Bool_0$.

Assuming Coquand or Russell universes and $p : i < j$, we can recover
polymorphic functions, for example, we may have $\mi{id} : \Pi(A : \U\,i\,j\,p)
(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity. Here,
we quantify over terms of $\U$, and since every type former stays on the same
level (including $\Pi$), we have to $\Lift$ the types in the codomain to the
level of the domain. We can also recover large elimination, for example as in
\[
(\lambda\,(b : \Bool_j).\,\mathsf{if}\,b\,\mathsf{then}\,\Code\,\top_i\,\mathsf{else}\,\Code\,\bot_i)
: \Tm_j\,\Gamma\,(\Bool_j \to \U\,i\,j\,p).
\]

@@

\begin{mydefinition} Fix a $\Lvl$ structure. A model for a type theory with $\Lvl$-indexed
universes consists of the following:
\begin{enumerate}
  \item A base category ($\Con$, $\Sub$) with a terminal object $\emptycon$.
  \item A strict family diagram ($\Ty_i$, $\Tm_i$) over $\Lvl$, supporting Russell
        universes, and each family structure is closed under the same basic type formers.
\end{enumerate}
\end{mydefinition}

We abbreviate the theory as TTGU, for ``type theory with generalized
universes''. The choice of available basic type formers is up to personal taste,
and it will not significantly affect the following model construction.

@@

IR codes, code lifting, properties,

\subsection{Model}

List a bunch of things in the model.

\section{Type Theories with First-Class Universe Levels}
\label{sec:ttfl}

>>>>

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\uir}{\msf{U^{IR}}}
\newcommand{\elir}{\msf{El^{IR}}}
\newcommand{\ult}{\U_{<}}

@@

\begin{enumerate}
\item In Section \ref{sec:ttgu} we describe models of type theories where
  universe levels may come from any set with a well-founded transitive ordering
  relation. We specify models as categories equipped with level-indexed diagrams
  of families, as a variation on categories with families. Each morphism of
  levels is mapped to a lifting operation on terms and types. By varying the
  preservation properties of lifting operations, we can describe a range of
  stratification features, from two-level type theory to cumulative universes.
\item We use induction-recursion to model the mentioned theories. We model
  the strongest formulations for lifting and universes, namely cumulative
  universes with Russell-style type decoding.
\item In Section \ref{sec:ttfl} we describe type theories with internal types
  for levels and level morphisms, and extend the previous inductive-recursive
  semantics to cover these as well. Here, we can additionally represent various
  universe polymorphism features and level computations. We illustrate how
  predicative universe features in Coq and Agda can be justified in our framework.
\end{enumerate}

@@

\section{Generalized Universe Hierarchies}
\label{sec:ttgu}

@@

\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$, all
descending $R$-chains must be finite. This is ensured by the universal property
of the inductive definition, as all inductive types must be well-founded.

\begin{mylemma}\label{lem:accprop}
Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
equal \cite[Lemma 10.3.4]{hottbook}.
\end{mylemma}

@@

Unlike every other type or term former, there is no lifting computation rule for
$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
lifting and term lifting, but while term lifting is invertible, type lifting is
not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a larger $\Ty$ is
not isomorphic a smaller one, because it contains more universes. So, for
example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$ yields
$\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.

Assuming Coquand or Russell universes and $p : i < j$, we can recover
polymorphic functions, for example, we may have $\mi{id} : \Pi(A : \U\,i\,j\,p)
(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity
function. Here, we quantify over terms of $\U$, and since every type former
stays on the same level (including $\Pi$), we have to $\Lift$ the types in the
codomain to match the level of the domain. We can also recover large
elimination, for example as in
\[
(\lambda\,(b : \Bool_j).\,\mathsf{if}\,b\,\mathsf{then}\,\Code\,\top_i\,\mathsf{else}\,\Code\,\bot_i)
: \Tm_j\,\Gamma\,(\Bool_j \to \U\,i\,j\,p).
\]

\todo{ordinal indexed levels and lubs}

@@

\begin{mydefinition}[Notion of model for a type theory with generalized universes (TTGU)]
Fix a $\Lvl$ structure. A model for TTGU consists of
\begin{enumerate}
  \item A base category ($\Con$, $\Sub$) with a terminal object $\emptycon$.
  \item A strict family diagram ($\Ty_i$, $\Tm_i$) over $\Lvl$, supporting Russell
        universes, and each family structure is closed under the same basic type formers.
\end{enumerate}
\end{mydefinition}

The choice of available basic type formers is up to personal taste, and it will
not significantly affect the following model construction.

Both in families and universes we choose the stricter formulation, since if we
give a model which proves the strict syntax consistent, we immediately get a model which
proves the weak syntax consistent\footnote{We always get \emph{initial} and \emph{terminal}
models automatically, because of the algebraic character of the theories in this paper. We
also get a \emph{freely generated} strict model from a weak model, from the left adjoint
of the functor which forgets the strictness equations. But none of these tricks can be used
to automatically get a consistency proof.}.

@@

The task is to interpret the $\Lvl$-many universes of TTGU using a metatheoretic
feature which is more compact and already well-studied. For this, we need to
define a $\Lvl$-indexed type of type codes. Since $\Lvl$ and $\!\blank<\!\blank$
can be arbitrary, we effectively need to define transfinite hierarchies of
codes. We use an inductive-recursive (IR) \cite{dybjer99finite} definition, for
the following reasons.

First, IR is already supported in the proof assistant Agda, and it is very
useful to be able to sketch out ideas in a machine-checked setting. It would be
much harder to do the same when developing semantics in set theory.

Second, could we use weaker type-theoretic features, like super universes
\cite{Palmgren98onuniverses} or Mahlo universes \cite{setzer00mahlo}? These are
sufficient to model transfinite hierarchies. However, it is not clear how to
additionally support \emph{recursive sub-universes}, which in our work corresponds to
the type former preservation property of $\Lift$. Thus, we give a custom IR
definition which corresponds more directly to TTGU structure.

\begin{mydefinition}[IR codes]
  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define
  $\uir$ and $\elir$ by simultaneous IR. For basic type formers, below we only
  include codes for function types and Booleans; other type formers are
  straightforward to add (and we do have more in our Agda formalization).
\begin{alignat*}{5}
  & \uir   &&: \Set0
     && \rlap{$\elir : \uir \to \Set0$} \\
  & \U'    &&: (j : \Lvl) \to j < i \to \uir
     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
  & \Bool' &&: \uir
     && \elir\,\Bool'      &&= \Bool
\end{alignat*}
\end{mydefinition}

\begin{notation}
We may write $\uir_{i\,f}$ and $\elir_{i\,f}$ in order to make parameters explicit.
\end{notation}

($\uir$, $\elir$) can be viewed as a \emph{universe operator}: given semantics
for an initial segment of $\Lvl$ (given by $i$ and $f$), we create a new
universe which is closed under basic type formers, and also closed under all
sets in $f$ by the way of $\U'$. Most importantly, this operation can be
transfinitely iterated. We first define universes for initial segments of
$\Lvl$, by induction on the accessibility of levels:
\begin{alignat*}{3}
  & \ult\,: \{i : \Lvl\}\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
  & \ult\,\{j\}\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})}
\end{alignat*}

\begin{mydefinition}[Semantic universe]
Since every level is accessible, we can define the full semantic hierarchy and
its decoding function.
\begin{alignat*}{3}
  &\U : \Lvl \to \Set0
     && \El : \{i : \Lvl\} \to \U\,i \to \Set0 \\
  &\U\,i = \uir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
     && \El\,\{i\} = \elir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}
\end{alignat*}
\end{mydefinition}

\begin{mylemma}\label{lem:ucomp}
Assuming $p : i < j$, we have the computation rule $\ult\,j\,p = \U\,j$. Proof: we
may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
some $f$. Then the equation becomes $\uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})} =
\uir_{j\,(\ult\,\{j\}\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
the $f\,j\,p$ and $\mathsf{<\!wf}\,j$ witnesses are equal. \qed
\end{mylemma}

\begin{mydefinition}[Semantic $\Lift$]
We define by induction on $\uir$ a function with type $(p : i < j)(A : \U\,i)
\to (A' : \U\,j)\times(\El\,A' = \El\,A)$. However, for the sake of clarity, we
present this here as two (mutual) functions:
\begin{alignat*}{3}
  &\Lift\,      &&: (p : i < j) \to \U\,i \to \U\,j\\
  &\msf{ElLift} &&: (p : i < j)(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
\end{alignat*}
Let us look at $\Lift$ first:
\begin{alignat*}{3}
  &\Lift\,p\,(\U'\,k\,q)    &&= \U'\,k\,(p \circ q)\\
  &\Lift\,p\,(\Pi'\,A\,B)   &&= \Pi'\,(\Lift\,p\,A)\,(\lambda\,a.\,\Lift\,p\,(B\,a))\\
  &\Lift\,p\,\Bool'         &&= \Bool'
\end{alignat*}
Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the
proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
show $\ult\,\{j\}\,k\,(p \circ q) = \ult\,\{i\}\,k\,q$, but by Lemma
\ref{lem:ucomp} both sides are $\U\,k$. \qed
\end{mydefinition}

\begin{mylemma} Properties of $\Lift$:
\begin{enumerate}
  \item $\Lift$ preserves all basic type formers; this is immediate from the definition.
  \item $\Lift$ is functorial, i.e.\ $\Lift\,(p \circ q)\,A = \Lift\,p\,(\Lift\,q\,A)$. This follows
    by induction on $A$, and we make use of the irrelevance of $\blank\!<\!\blank$ in the $\U'$ case.
    \qed
\end{enumerate}
\end{mylemma}

\subsection{Model of TTGU}

We give a model of TTGU in this section.

\todo{disambiguate IR and TTGU type notation}

\begin{mydefinition}[Base category]
This is simply the category of sets and functions in $\Set0$, i.e.\ $\Con = \Set0$,
$\Sub\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
\end{mydefinition}

\begin{mydefinition}[Family diagram]
We map $i : \Lvl$ to a family structure as follows.
\begin{alignat*}{5}
  &\Ty_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \Tm_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
\end{alignat*}
Type and term substitution are given by composition with some $\sigma : \Gamma
\to \Delta$ function. Context extension is given by $\Gamma \ext A = (\gamma :
\Gamma) \times \El\,(A\,\gamma)$. Since we define $\blank\!\ext\!\blank$ as a
$\Sigma$-type, the comprehension structure for it follows immediately, as
$\Sigma$ pairing and projections.

Mapping $i$ to a strict inclusion is defined as follows. We
\end{mydefinition}

\section{First-Class Universe Levels}
\label{sec:ttfl}
